The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-02-22 paper.tex
LLMs and direct manipulation

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\kvd}[1]{\textsf{\bfseries #1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\suggest}{\textnormal{\ident{suggest}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
%\newcommand{\preview}{\textnormal{\ident{preview}}}
%\newcommand{\links}{\textnormal{\ident{links}}}
\DeclareMathOperator{\argmax}{arg\,max}

@@

\begin{abstract}

Some of the most remarkable results in mathematics reveal connections between different
branches of the discipline. The aim of this paper is to point out a modest, but still
remarkable, similarity between a range of different interactive programming systems.

@@

use a simple formal mathematical model
that we call \emph{the choose-your-own-adventure calculus} to

TODO: Say somewhere that this also formalizes things like mixed-initiative interaction and
programming by demonstration

% and enable a transfer of ideas between them

@@

When the AI assistants runs automatically, it correctly maps the numerical columns, but it
incorrectly maps the \ident{Urban.rural} (2014) column to \ident{Nation} (2015). This happens
because both columns are categorical and have three values with similar distribution. A data
analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
\ident{Don't match Urban.rural and Nation} to specify that the two columns should not be matched.
Datadiff then runs again and finds the correct matching.

@@

\begin{itemize}
\item A system for type-directed program synthesis could automatically synthesize parts of a
  program (e.g., pattern matching and implementation for some of the cases), but ask user for
  input in order to complete branches where solution was not found automatically.
\item In data exploration, a system could perform automatic search through the available operations
  in order to transform data into a more regular format, according to a suitability
  metric as done, for example, in the Proactive Wrangling system~\cite{guo-2011-proactive}.
\end{itemize}

@@

\subsection{LLM assistant for data exploration}

\subsection{Programming by demonstration}

ways of integrating automation with manual interaction
- confirm each step vs. auto and retrace

\cite{rein-2019-exploratory}

\cite{verter-2024-mixed}

@@

hard to edit - you cannot easily go back

@@

% cite: https://link.springer.com/chapter/10.1007/3-540-63104-6_39

>>>>

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\kvd}[1]{\textsf{\bfseries #1}}
\newcommand{\ddident}[1]{\guillemotleft\ident{#1}\guillemotright}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\suggest}{\textnormal{\ident{suggest}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\vect}[1]{\lbrack #1 \rbrack}
\newcommand{\preview}{\textnormal{\ident{preview}}}
\newcommand{\links}{\textnormal{\ident{links}}}
\DeclareMathOperator{\argmax}{arg\,max}

@@

\begin{abstract}
\noindent
TODO: Some of the most remarkable results in mathematics reveal connections between different
branches of the discipline. The aim of this paper is to point out a modest, but still
remarkable, similarity between a range of different interactive programming systems.

@@

\noindent
TODO: Say somewhere that this also formalizes things like mixed-initiative interaction and
programming by demonstration

\noindent
TODO: It is mainly about giving us a nice way to talk about lots of things in interactive
programming systems and enable a transfer of ideas between them

@@

When the AI assistants runs automatically, it correctly maps the numerical columns, but it
incorrectly maps the \ident{Urban.rural} (2014) column to \ident{Nation} (2015). This happens
because both columns are categorical and have three values with similar distribution. A data
analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
\ident{Don't match `Urban.rural' and `Nation'} to specify that the two columns should not be matched.
Datadiff then runs again and finds the correct matching.

@@

\begin{itemize}
\setlength{\itemsep}{5pt}
\item A system for type-directed program synthesis could automatically synthesize parts of a
  program (e.g., pattern matching and implementation for some of the cases), but ask user for
  input in order to complete branches where solution was not found automatically.
\item In data exploration, a system could perform automatic search through the available operations
  in order to transform data into a more regular format, according to a suitability
  metric as done, for example, in the Proactive Wrangling system~\cite{guo-2011-proactive}.
\end{itemize}

@@

\subsection{Language model-based completion}
Large language models can assist with data exploration by generating snippets of code based
on natural language description of the problem \cite{yin-2023-codegen}. A recognized drawback of
this approach is that the user may gain only cursory understanding of the code and overlook
errors. Various systems address this by generating explanations alongside with code \cite{nooralahzadeh-2024-data}.
We developed a system that provides LLM-based natural language assistance for The Gamma [report
citation omitted] based on a mechanism that assists the user and is also applicable to other
choose-your-own-adventure systems.

\begin{figure}
{\raggedright\ttfamily\small
You are helping user to complete a task in an interactive programming environment. The user's
query is: "Give me the athlete with the largest number of gold medals."

~\\[-0.75em]

The query built so far is: "olympics"."group data"."by Athlete".

~\\[-0.75em]

The environment offers the user possible options. Choose an option that the should be applied to the current dataset:

~\\[-0.75em]

1.\hspace{0.4em} count distinct Athlete\\
2.\hspace{0.4em} count distinct Discipline\\
\hspace{2.3em}\lbrack\textit{multiple options omitted}\rbrack \\
13. count all\\
14. concatenate Athlete\\
15. concatenate Discipline\\
\hspace{2.3em}\lbrack\textit{multiple options omitted}\rbrack\\
23. sum Gold\\
24. sum Silver\\
25. sum Year\\

~\\[-0.75em]

You should answer with the number of the option and no further explanation.}
\caption{A prompt to complete a data exploration query based on a natural language
question, which uses LLM to choose from the available completion options.}
\label{fig:llm}
\end{figure}

\subparagraph{LLM assistant for The Gamma.}
Our system [reprot citation omitted] integrates a type provider like the one used in The Gamma
with a large language model (GPT-3.5 and GPT-4). It lets user ask a natural language query and
then recommends choices that the user should make in order to answer the query. Our system does
not use LLM to generate code, but to recommend an option offered by the type provider. To do so,
we construct a prompt as shown in Figure~\ref{fig:llm}, which asks the LLM to recommend the next
choice. In this example, the LLM is able to reliably respond with ``23'', which is the correct
choice. Our system then pre-selects it in the completion list offered to the user.

In contrast to one-shot generation of Python or SQL code snippets, our approach guides the user
through the program construction, providing them with an opportunity to review and check if the
program logic matches their expectations.

In addition to the basic prompting strategy outlined above, we evaluated a strategy where the LLM
is provided with lookahead information, i.e., for each of the choices, we also include a list of
the choices that will available subsequently (formatted as either inline information in parentheses
or as a nested bullet-point list). We use two different data sources to evaluate the system,
each with 10 queries for which we manually determine the correct choices. To compare the results,
we compute a score for each query by following the correct path, asking for an LLM recommendation
in each step and computing the ratio of correct choices. Figure~\ref{fig:methods} shows the
average scores for five different strategies, using two different LLM engines.

\begin{figure}[t]
\includegraphics[width=0.7\textwidth]{fig/llm.png}
\caption{A plot showing the average scores for two different large language models (GPT-3.5 and GPT-4)
solving 20 challenge problems using five different prompting strategies.}
\label{fig:methods}
\end{figure}

The results suggest that, when using a more advanced LLM system, the ratio of correct
recommendations is high-enough to be practically useful. It also shows that providing more
information to the LLM through the lookahead mechanism improves the quality of the recommendation,
in a consistent way across multiple LLM systems. Arguably, offering a recommendation is preferrable
over providing an opaque solution, because it keeps the human in the loop \cite{sellen-2024-copilot}
and avoids ``ironies of automation'' \cite{bainbridge-1983-ironies} including deskilling.

\subparagraph{Generalised LLM assistant.}
The LLM-based recommendation engine developed for The Gamma can be implemented for any
choose-your-own-adventure system. As can be seen in Figure~\ref{fig:llm}, the information needed
to construct an LLM prompt comprise only the natural language query entered by the user,
a sequence of previously made choices $\iota_1, \ldots, \iota_{n-1}$ and the identifiers
$\iota_n, \iota_n', \iota_n'', \ldots$ of the choices offered by the $\choices$ operation.
As the LLM-based recommendations are based on natural language analysis of the prompt,
the quality of the recommendations depends on how semantically meaningful the generated
identifiers are.

An LLM-based recommendation engine can be useful for a number of the choose-your-own-adventure
interactive systems discussed in this paper:

\begin{itemize}
\setlength{\itemsep}{5pt}
\item We demonstrated the usefulness of the system for data exploration in The Gamma.
  Type providers for structured data \cite{petricek-2016-fsdata} typically generate small number
  of choices that the user can navigate without assistance, but navigating the schemas generated by
  type providers for semantic knowledge bases \cite{syme-2012-inforich} may be simplified through a
  natural language query.
\item The datadiff AI assistant discussed in Section~\ref{sec:motivation} matches columns based
  on statistical distribution and ignores column names. An LLM-based recommendation engine is
  useful in this case, because it is able to recommend the option \ident{Don't match
  `Urban.rural' and `Nation'} based solely on its name. For other AI assistants, the LLM-based
  recommendation engine would need access to the input data in order to be useful.
\item The problem of predicting steps in order to construct a proof is known as
  proofstep generation in literature focused on deep learning for theorem
  proving~\cite{zhaoyu-2024-deep}. Although most systems use models trained specifically for the
  task, there is also interest in using general-purpose LLMs~\cite{zhang-2023-proofs}.
  Our approach suggests a potential prompting strategy.
\end{itemize}

It is interesting to note that the choose-your-own-adventure calculus provides a common structure
that makes it possible to integrate multiple different approaches to assisting human in program
construction, ranging from automatic search (if we can provide a suitable heuristic) to
the use of LLMs (if generated choices carry enough semantic information). It can also serve as the
basis for visual programming environments discussed next.

\begin{figure}[t]
  \includegraphics[width=0.9\textwidth]{fig/histogram.png}
  \caption{Programming by demonstration in Histogram -- (1) the user constructs program to load
    data, (2) then they filter data using graphical interface, which (3) records an operation
    corresponding to the interaction in code.}
  \label{fig:histogram}
\end{figure}

\subsection{Direct manipulation}

The Histogram programming environment~\cite{petricek-2019-histogram} provides a code completion
mechanism that can be used both to choose operations (as with type providers) and to specify
their parameters (by offering available values in context as possible arguments). It makes it
possible to evaluate sub-expressions at a fine-grained level and also refines type information
based on runtime values. However, it also supports programming by demonstration
\cite{cypher-1993-pbd,kandel-2011-wrangler}.

As illustrated in Figure~\ref{fig:histogram}, when the user loads data, they can manipulate it
in a table through a direct manipulation interface~\cite{shneiderman-1983-direct}. In Histogram,
interacting with the table triggers a sequence of operations that construct code. A similar
functionality is available in The Gamma, where operations for manipulating tabular data can be
selected not only through auto-completion, but also by interacting with the live preview.

The choose-your-own-adventure calculus enables a more general perspective on the programming
by demonstration interaction as implemented by Histogram. When interacting with a live
preview, the user is using a graphical interface to choose an option offered by $\choices$.
The operation appends an item to the sequence of choices, updates the system state
and generates a new preview where the choice is reflected (e.g., by filtering the table).

An interesting question is whether a program can be constructed solely through direct manipulation,
by interacting with the live previews. For this to be possible, the live preview needs to
offer links to all the possible options offered by $\choices$. This requires a careful design
of live previews. In Histogram (Figure~\ref{fig:histogram}), the live preview offers interactive
elements for filtering based on equality test (row of drop-downs), sorting (up/down arrows)
and indexing (indices in the ``\#'' column), but some operations (such as aggregation) can only
be constructed through code. Previews in The Gamma provide interactive elements for all options,
but those cannot always be integrated with the data display. For example, after choosing a grouping
key (Figure~\ref{fig:groupby}), the preview shows the key column, but other columns are hidden.
Specifying an aggregation thus requires an additional (somewhat arbitrary) ``+'' button that
provides access to the available choices.

\begin{figure}[t]
\includegraphics[width=0.6\textwidth]{fig/thegamma3.png}
\caption{Specifying aggregation in The Gamma through a live preview. Grouping keys can be
  selected from a drop-down, while aggregations have to be added through a menu.}
\label{fig:groupby}
\end{figure}

We can use the choose-your-own-adventure calculus to make the question whether all
programs can be constructed through direct manipulation more precise. Assume
$\preview(e)=p$ is a preview constructed for an expression $e$ and $\links(p)=\iota_1, \ldots, \iota_n$
are identifiers that can be invoked through the preview (akin to links in a hypertext document).

\begin{definition}[Direct manipulation]
A choose-your-own-adventure system with previews defined by $\preview$ and $\links$ supports
full direct manipulation if:
\begin{itemize}
\item $\forall \sigma, e$ such that $\select(\sigma)=e$ it is the case that
  $\forall \iota'\mapsto\sigma'\in\choices(\sigma)\,.\,\iota'\!\in\!\links(\preview(e))$.
\end{itemize}
\end{definition}

A system supports full direct manipulation if any program can be constructed by interacting with
the live previews, created based on the gradually constructed programs. As illustrated by The Gamma,
this can always be achieved by listing the options in a menu, but it is more interesting to see
if the links can be directly embedded in the preview as in the data table interface of Histogram.

The idea of directly mapping elements in a user interface to underlying structure of the
programming system has previously been implemented in pioneering user interface systems
including the Alternate Reality Kit~\cite{randall-1986-ark} and the Morphic framework for
Self and Squeak \cite{maloney-1995-morphic,maloney-2001-morphic}. In those systems, user interface
interactions directly map to messages sent to the underlying object and the halo element in
Morphic plays a similar role to the additional menu in The Gamma. It remains to be seen if
the choose-your-own-adventure calculus can provide a new way of looking at those systems.

@@

TODO: Is this too general to be practically useful? E.g. LLM for interactive theorem provers
arguably needs more specific information.

\noindent
TODO: What if the list of choices is infinite?

\noindent
TODO: How do you go back and revisit earlier choices? Backtracking in theorem provers or
fixing earlier part of the query in data exploration. Sometimes, you can retrace the
steps, but they may change...

\noindent
Related work - mbark for emacs, VS code command palette

\noindent
Alphageometry paper - follows a similar way of working to choose-your-own-adventure

\section{Conclusions}

TODO: Some conclusions
1) lets us ask all those questions
2) let ideas cross between domains

@@
